Testing Quasar Unification: I. Radiative Transfer in Clumpy Winds

### 2016-01-26 paper.tex
mod2, sent around

\maketitle
\begin{abstract} 
Various unification schemes interpret the complex phenomenology of quasars and luminous active galactic nuclei (AGN) in terms of a simple picture involving a central black hole, an accretion disc and an associated outflow. Here, we continue our tests of this paradigm by comparing quasar spectra to synthetic spectra of biconical disc wind models, produced with our state-of-the-art Monte Carlo radiative transfer code. Previously, we have shown that we could produce synthetic spectra resembling those of observed broad absorption line (BAL) quasars, but only if the X-ray luminosity was limited to
10$^{43}$~erg~s$^{-1}$. Here, we show that clumping the outflow can maintain the ionization state necessary for BAL features in the rest-frame UV at more realistic X-ray luminosities. Our fiducial model shows good agreement with X-ray properties of AGN and the dense, X-ray heated wind produces strong line emission in, e.g., \la\ and \civline at low inclinations. At high inclinations, the synthetic spectra possess prominent LoBAL features. Despite these successes, we cannot reproduce all the emission lines seen in quasar spectra with the correct equivalent-width ratios, and and we find an angular dependence of emission-line equivalent width despite the remarkable similarities in the observed emission line properties of BAL and non-BAL quasars. Overall, our work suggests that a biconical wind can reproduce much of the qualitative behaviour expected from a unified model, but we cannot yet provide quantitative matches with quasar properties at all viewing angles. Whether disc winds can successfully unify quasars is therefore still an open question.
\end{abstract}

@@

Despite the importance of disc winds in shaping quasar and AGN spectra,  
much of the underlying outflow physics remains uncertain. 
Several driving mechanisms have been proposed, including
thermal pressure \citep{weymann1982, begelman1991}, magnetocentrifugal forces 
\citep{blandfordpayne,pelletier_pudritz} and 
radiation pressure on spectral lines \citep[`line-driving';][]{lucysolomon1970,shlosman1985,MCGV95}.
Of these, line-driving is possibly the most attractive, as
strong absorption lines are already seen in BALQSOs and the X-ray spectra of AGN 
\citep{reeves2003,poundsreeves2009,tombesi2010a}.
The efficiency of line-driving is crucially dependent on the ionization state 
of the outflowing plasma, meaning that it is difficult to prevent 
the wind becoming over-ionized and `failing' in the presence of strong X-rays. 
\cite{MCGV95} proposed a potential solution: 
a region of `hitchhiking gas' that could shield the wind from the central X-ray source. 
An additional or alternative solution is that the wind is clumped 
\citep[e.g.][]{hamann2013}
possibly on multiple scale lengths. Local density enhancements could lower the 
ionization parameter of the plasma while still maintaining the same mass-loss 
rate and column density.

Evidence for dense substructures in AGN winds is widespread.
BALQSOs show complex absorption line profiles \citep{ganguly2006, simonhamann2010}
and exhibit variability in these profile shapes \citep{capellupo2011,capellupo2012,capellupo2014}.
AGN generally show variability in X-ray absorption components \citep[e.g.][]{risaliti2002}
and many models for the BLR consist of clumps embedded in an outflow 
\citep{krolik1981, emmering1992, dekool1995, cassidyraine1996}.
Clumping can be caused by magnetic confinement \citep{dekool1995},
or the instabilities inherent to line-driven winds 
\citep{lucysolomon1970,macgregor1979,carlberg1980,owockirybicki1984,owockirybicki1985}.
Additionally, clumping is required to explain the electron scattering wings of emission lines formed
in line-driven hot star winds \citep{hillier1991eswingsmodel}. Complex substructures 
are also produced in simulations of line-driven 
outflows in AGN, although on very different scales to line-driven instabilities 
\citep{PSK2000,PK04,progakurosawa2010,proga2014}.  
In summary, clumpy winds offer an observationally motivated and theoretically 
predicted way to lower the ionization state of a plasma, possibly in tandem
with a shielding scenario.

@@

We include  free-free, bound-free and bound-bound heating
and cooling processes in the model. For radiative transfer purposes
we treat electron scattering in the Thomson limit, 
but take full account of Compton heating and cooling when
calculating the thermal balance of the plasma (see H13).
{\bf 
Adiabatic cooling in the cell is calculated using the estimator
\begin{equation}
C_{ad} = f_V n_e k_B T_e V (\Delta \cdot v),
\end{equation}
where $V$ is the cell volume, $T_e$ is the electron temperature,
$n_e$ is the electron density and $\Delta \cdot v$ is the divergence of the 
velocity field at the centre of the cell.
For self-consistency, $k$-packets then have a finite 
destruction probability of 
\begin{equation}
P_{k,destruct} = \frac{C_{ad}}{C_{tot}},
\end{equation}
where $C_{tot}$ is the sum over all cooling rates in the cell.
We find that adiabatic cooling is insignificant in much of
the outflow.

@@

In the SV93 model, a biconical disc wind rises from the accretion 
disc between launch radii $r_{min}$ and $r_{max}$.
The opening angles of the wind are set to $\theta_{min}$ and $\theta_{max}$.
The poloidal velocity along each individual streamline at a poloidal distance $l$ 
is then given by
\begin{equation}
v_l=v_0+\left[v_{\infty}(r_0)-v_0\right]\frac{\left(l/R_v\right)^{\alpha}}{\left(l/R_v\right)^{\alpha}+1},
\label{v_law}
\end{equation}
where $v_0$ is the velocity at the base of the streamline, $\alpha$ is
an exponent governing how quickly the wind accelerations and 
$R_v$ is the `acceleration length', defined as the distance at which
the outflow reaches half of its terminal velocity, $v_{\infty}$.
The terminal velocity is set to a fixed multiple of the escape
velocity, $v_{esc}$, at the base of the streamline (radius $r_0$).
The rotational velocity, $v_{\phi}$, is initially Keplerian ($v_k = [GM/r_0]^{1/2}$),
and the wind conserves specific angular momentum, such that 
\begin{equation}
v_{\phi} r = v_k r_0.
\label{v_law}
\end{equation}
The velocity law is crucial in determining the output spectra,
as it affects not only the projected velocities along the line of sight,
but also the density and ionization state of the outflow.
A wind that accelerates more slowly will have a denser wind base
with correspondingly different ionization and emission characteristics.
}

@@

In our previous modelling efforts, we assumed a smooth outflow, 
in which the density at a given point was determined only by the 
kinematic parameters and mass loss rate. However, as already discussed,
AGN winds exhibit significant substructure -- the outflow is expected to be
{\em clumpy}, rather than smooth, and probably on a variety of scales. 
A clumpy outflow offers a possible solution to the so-called `over-ionization problem' in 
quasar and AGN outflows \citep{hamann2013}. This is the main motivation for incorporating 
clumping into our model.

Implementing a treatment of clumping is challenging, for
two main reasons. First, the physical scale lengths and density contrasts 
in AGN outflows are not well-constrained from observations.  
Second, there are significant computational difficulties associated with adequately
resolving and realistically modelling a series of small scale, high density
regions with a MCRT code. 
Given the lack of knowledge about the actual type of clumping, we have adopted
a simple approximation
used successfully in stellar wind modelling, known as 
{\em microclumping} \citep[e.g.][]{hamann1998,hilliermiller1999,hamann2008}. 
The underlying assumption of microclumping is that clump sizes are much smaller than the 
typical photon mean free path, and thus the clumps are 
both geometrically and optically thin. This approach 
allows one to introduce a `volume filling factor', $f_V$. 
The intra-clump medium is assumed to be a vacuum, so the 
density of the clumps is then multiplied by the ``density enhancement'' 
$D=1/f_V$. Opacities, $\kappa$, and emissivities, $\epsilon$, 
can then be expressed as 
\begin{equation}
\kappa = f_V \kappa_C(D);~~\epsilon = f_V \epsilon_C(D).
\end{equation}
Here the subscript $C$ denotes that the quantity is calculated using the 
enhanced density in the clump. The resultant effect is that, {\em for fixed temperature},
 processes that are linear in density, such as electron scattering, are unchanged, 
as $f_V$ and $D$ will cancel out. However, any quantity that scales with the square of density, 
such as collisional excitation or recombination, will increase by a factor of $D$.

In our models, the temperature is not fixed, and is instead set by balancing heating and 
cooling in a given cell. In the presence of an X-ray source, this thermal balance is 
generally dominated by bound-free heating and line cooling. The main effect of including 
clumping in our modelling is that it moderates the ionization state due to the increased 
density. This allows an increase in the ionizing luminosity, amplifying the amount of
bound-free heating and also increasing the competing line cooling term
(thermal line emission).
Our clumping treatment is necessarily simple; it does not adequately
represent the complex substructures and stratifications in ionization
state we expect in AGN outflows. Nevertheless, this parameterization allows simple estimates
of the effect clumping has on the ionization state and emergent 
line emission.

@@

{\bf This extreme $89^\circ$ viewing angle cannot represent a typical quasar within a unified model, as
extreme inclinations should be extremely underrepresented in quasar samples due to the
compton-thick absorber or `Torus' expected at high inclinations ().}
However, this spectrum does show that a model such as this can 
naturally reproduce quasar emission lines if the emissivity of the wind is 
increased {\em with respect to the disc continuum}.
In addition, it neatly demonstrates how a stratified outflow can naturally
reproduce the range of ionization states seen in quasars.

@@

At high inclinations, the synthetic spectra exhibit blue-shifted BALs in \al\ and \mg --
the absorption lines seen in LoBALQSOs, and we even see absorption in Fe~\textsc{ii}
at the highest inclinations. Line profiles in velocity space 
for \civ, \al\ and \mg, are shown in Fig.~\ref{fig:lobal} for a range
of BALQSO viewing angles. We find that ionization stratification
of the wind causes lower ionization material to have a smaller covering factor, 
as demonstrated by figures~\ref{fig:wind} and \ref{fig:lobal}.
This confirms the behaviour expected from a unification model such as Elvis (2000). 
LoBALs are only present at viewing angles close to edge-on ($i>75^\circ$),
as predicted by polarisation results \citep{brotherton1997}.
As observed in a BALQSO sample by \cite{filizak2014}, we find that
BAL troughs are wider and deeper when low ionization absorption features are present,
and high ionization lines have higher blue-edge velocities than the 
low ionization species.
There is also a correlation between the strength of LoBAL features
and the amount of continuum attenuation at that sightline, particularly
blueward of the Lyman edge as the low ionization base 
intersects the line-of-sight. 
A model such as this therefore predicts that LoBALQSOs and FeLoBALQSOs 
have stronger Lyman edge absorption and 
are more Compton-thick than HiBALQSOs and Type 1 quasars.
An edge-on scenario also offers a potential explanation for the rarity of LoBAL and
FeLoBAL quasars, due to a foreshortened and attenuated continuum, 
although BAL fraction inferences are fraught with complex selection 
effects \citep{goodrich1997,krolikvoit1998}.

@@

Having selected an individual fiducial model from the simulation grid, it is important
to briefly explore how specialised this model is, and how small parameter
changes can affect the synthetic spectra. Fig.~\ref{fig:grid}
shows the EW at a low inclination, and $BI$ at a high inclination for the simulation
grid. A few conclusions can be drawn from this plot straight away. 
First, we find that almost all the models with $f_V=0.1$ are over-ionized, and 
fail to produce strong \civ\ BALs or emission lines. Second, we find that 
it is difficult to significantly increase line emission while
keeping the luminosity and mass loss rate of the system fixed. 
We show an additional point on figure 7 corresponding to a model with an order of
magnitude higher X-ray luminosity and double the mass loss rate. As expected, 
this results in far higher line EWs, but fails to produce BALs because
the collisionally excited emission swamps the BAL profile. In addition,
this model would lie well above the expected $L_{2kev}-L_{2500}$ 
relation in figure 5. Such a high X-ray luminosity could therefore 
not be the cause of the strong line emission seen in {\em all} Type 1 quasars.

@@

We have carried out MCRT simulations using a simple
prescription for a biconical disc wind, with
the aim of expanding on the work of H13. To do this, we introduced two main
improvements: First, we included a simple treatment of clumping, and second, 
we improved the modelling of recombination lines by treating H and He as
`macro-atoms'. Having selected a fiducial model from an initial simulation grid,
we assessed the viability of such a model for geometric 
unification of quasars, and found the following main points:
\begin{enumerate}
\item Clumping moderates the ionization state
sufficiently to allow for the 
formation of strong UV BALs while agreeing well with the X-ray
properties of luminous AGN and quasars. 
\smallskip
\item A clumpy outflow model naturally 
reproduces the range of ionization states
expected in quasars, due to its stratified density
and temperature structure. 
LoBAL line profiles are seen at a subset of viewing angles, and Fe~\textsc{ii}
absorption is seen at particularly high inclinations. 
\smallskip
\item The synthetic spectra show \la\ line and weak He~\textsc{ii}~$1640$\AA\ line
as a result of our improved treatment of recombination using macro-atoms. We also see
Balmer emission lines and a Balmer recombination continuum in the optical spectrum, but this
is only really significant at high inclination where to continuum is suppressed.  
\smallskip
\item The higher X-ray luminosity causes a significant 
increase in the strength of the collisionally excited emission
lines produced by the model. 
However, the equivalent-width ratios of the emission lines do not match
observations, suggesting that a greater volume of dense ($n_e\sim10^{10}$~cm$^{-3}$)
material may be required.
\smallskip
\item The line EWs in the synthetic spectra increase with inclination.
BAL and non-BAL quasar composites have comparable EWs, so our model
fails to reproduce this behaviour.
 If the BLR emits fairly isotropically then for a 
foreshortened, limb-darkened accretion disc 
it is not possible to achieve line ratios at low inclinations 
that are comparable to those at high inclinations. 
We suggest that understanding the angular distribution of 
line and continuum emission is a crucial question for theoretical models.
\end{enumerate}
Our work confirms a number of expected outcomes from a geometric unification 
model, and suggests that a simple biconical geometry such as this can come close to 
explaining much of the  phenomenology of quasars. However, our conclusions pose 
some challenges to a picture in which BALQSOs are
explained by an {\em equatorial} wind rising from a classical thin disc, and suggest 
the angular distribution of emission is important to understand if this 
geometry is to be refuted or confirmed. We suggest that obtaining reliable 
observational orientation indicators and 
exploring a wider parameter space of outflow geometries in simulations
are obvious avenues for future work.

@@

The work of JHM, SWM, NSH and CK is supported by the
Science and Technology Facilities Council (STFC),
via two studentships and a consolidated grant, respectively.
CK also acknowledges a Leverhulme fellowship.
{\bf We would like to thank the anonymous referee for an insightful
and constructive report.}
We would like to thank Omer Blaes, Ivan Hubeny and Shane Davis for their
assistance with \agn. We are grateful to Mike Brotherton, Mike DiPompeo,
Sebastien Hoenig and Frederic Marin for helpful correspondence regarding
polarisation measurements and orientation indicators.
We would also like to thank Daniel Proga, Daniel Capellupo, Sam Connolly and
Dirk Grupe for useful discussions.  Simulations were conducted using \py\ version 80,
and made use of the IRIDIS High Performance Computing Facility at the
University of Southampton. Figures were produced using the {\tt matplotlib} 
plotting library \citep{matplotlib}. This work made use of the Sloan Digital Sky Survey.
Funding for the Sloan Digital Sky Survey has been provided by
the Alfred P. Sloan Foundation, the U.S. Department of Energy Office of
Science, and the Participating Institutions.

>>>>

\maketitle
\begin{abstract} 
Various unification schemes interpret the complex phenomenology of quasars and 
luminous active galactic nuclei (AGN) in terms of a simple picture involving a 
central black hole, an accretion disc and an associated outflow. Here, we continue
our tests of this paradigm by comparing quasar spectra to synthetic spectra of 
biconical disc wind models, produced with our state-of-the-art Monte Carlo radiative 
transfer code. Previously, we have shown that we could produce synthetic spectra 
resembling those of observed broad absorption line (BAL) quasars, but only if the 
X-ray luminosity was limited to 10$^{43}$~erg~s$^{-1}$. 
{\bf Here, we introduce a simple treatment of clumping, and find that a volume filling factor of $\sim0.01$ 
moderates the ionization state sufficently for BAL features to form in the 
rest-frame UV at more realistic X-ray luminosities.}
Our fiducial model shows good agreement with X-ray properties of AGN and the dense, 
X-ray heated wind produces strong line emission in, e.g., \la\ and \civline\ at low inclinations. 
At high inclinations, the synthetic spectra possess prominent LoBAL features. 
Despite these successes, we cannot reproduce all the emission lines seen in quasar 
spectra with the correct equivalent-width ratios, and we find an angular dependence of 
emission-line equivalent width despite the remarkable similarities in the observed emission 
line properties of BAL and non-BAL quasars. Overall, our work suggests that a biconical wind 
can reproduce much of the qualitative behaviour expected from a unified model, but we cannot 
yet provide quantitative matches with quasar properties at all viewing angles. Whether disc 
winds can successfully unify quasars is therefore still an open question.
\end{abstract}

@@

Despite the importance of disc winds in shaping quasar and AGN spectra,  
much of the underlying outflow physics remains uncertain. 
Several driving mechanisms have been proposed, including
thermal pressure \citep{weymann1982, begelman1991}, magnetocentrifugal forces 
\citep{blandfordpayne,pelletier_pudritz} and 
radiation pressure on spectral lines \citep[`line-driving';][]{lucysolomon1970,shlosman1985,MCGV95}.
Of these, line-driving is possibly the most attractive, as
strong absorption lines are already seen in BALQSOs and the X-ray spectra of AGN 
\citep{reeves2003,poundsreeves2009,tombesi2010a}.
The efficiency of line-driving is crucially dependent on the ionization state 
of the outflow, and in AGN, unlike O stars, 
the presence of strong X-ray emission can overionize the wind, 
causing it to `fail'.
\cite{MCGV95} proposed a potential solution: 
a region of `hitchhiking gas' that could shield the wind from the central X-ray source. 
An additional or alternative solution is that the wind is clumped 
{\bf \citep[e.g.][]{junk1983,weymann1985,hamann2013}}
possibly on multiple scale lengths. Local density enhancements could lower the 
ionization parameter of the plasma 
without requiring excessively large total column densities 
and mass-loss rates.

Evidence for dense substructures in AGN winds is widespread.
BALQSOs show complex absorption line profiles \citep{ganguly2006, simonhamann2010}
and exhibit variability in these profile shapes \citep{capellupo2011,capellupo2012,capellupo2014}.
AGN generally show variability in X-ray absorption components \citep[e.g.][]{risaliti2002}
and many models for the BLR consist of clumps embedded in an outflow 
\citep{krolik1981, emmering1992, dekool1995, cassidyraine1996}.
{\bf 
\cite{krolik1981} showed that BLR clouds would be short-lived, suggesting 
that a confining mechanism is required, such as
magnetic confinement \citep[e.g.][]{dekool1995}. Alternatively,
the `line deshadowing instability' (LDI) inherent in line-driven winds
\citep{lucysolomon1970,macgregor1979,carlberg1980,owockirybicki1984}
may be responsible for smaller-scale clumping, as is observed in O-star winds
\citep[][and references therein]{fullerton2011}.
Indeed, simple models of clumping have successfully explained the electron scattering 
wings of emission lines formed in these line-driven flows 
\citep{hillier1991eswingsmodel}.}
Complex substructures are also produced in simulations of line-driven 
outflows in AGN, although on very different scales to LDIs 
\citep{PSK2000,PK04,progakurosawa2010,proga2014}.  
In summary, clumpy winds offer an observationally motivated and theoretically 
predicted way to lower the ionization state of a plasma, possibly in tandem
with a shielding scenario. However, it is equally important to note that there 
is no consensus view of the size or location of the clumps in the outflow.

@@

We include  free-free, bound-free and bound-bound heating
and cooling processes in the model. For radiative transfer purposes
we treat electron scattering in the Thomson limit, 
but take full account of Compton heating and cooling when
calculating the thermal balance of the plasma (see H13).
{\bf Adiabatic cooling in the cell is calculated using the estimator
\begin{equation}
C_{A} = f_V k_B T_e V (\nabla \cdot v) \left(n_e + \sum\limits_{i=1}^{ions} N_i\right),
\end{equation}
where $V$ is the cell volume, $T_e$ is the electron temperature,
$n_e$ is the electron density, $N_j$ is the number density of ion $j$
and $\nabla \cdot v$ is the divergence of the 
velocity field at the centre of the cell. The sum is over all ions 
included in the simulation.
For consistency, $k$-packets then have a finite 
destruction probability of 
\begin{equation}
P_{k,A} = \frac{C_{A}}{C_{tot}},
\end{equation}
where $C_{tot}$ is the sum over all cooling rates in the cell.
}

@@

In the SV93 model, a biconical disc wind rises from the accretion 
disc between launch radii $r_{min}$ and $r_{max}$.
The opening angles of the wind are set to $\theta_{min}$ and $\theta_{max}$.
The poloidal velocity along each individual streamline at a poloidal distance $l$ 
is then given by
\begin{equation}
v_l=v_0+\left[v_{\infty}(r_0)-v_0\right]\frac{\left(l/R_v\right)^{\alpha}}{\left(l/R_v\right)^{\alpha}+1},
\label{v_law}
\end{equation}
where $v_0$ is the velocity at the base of the streamline, $\alpha$ is
an exponent governing how quickly the wind accelerations and 
$R_v$ is the `acceleration length', defined as the distance at which
the outflow reaches half of its terminal velocity, $v_{\infty}$.
The terminal velocity is set to a fixed multiple of the escape
velocity, $v_{esc}$, at the base of the streamline (radius $r_0$).
The rotational velocity, $v_{\phi}$, is initially Keplerian ($v_k = [GM/r_0]^{1/2}$),
and the wind conserves specific angular momentum, such that 
\begin{equation}
v_{\phi} r = v_k r_0.
\label{v_law}
\end{equation}
The velocity law is crucial in determining the output spectra,
as it affects not only the projected velocities along the line of sight,
but also the density and ionization state of the outflow.
A wind that accelerates more slowly will have a denser wind base
with correspondingly different ionization and emission characteristics.

@@

In our previous modelling efforts, we assumed a smooth outflow, 
in which the density at a given point was determined only by the 
kinematic parameters and mass loss rate. However, as already discussed,
AGN winds exhibit significant substructure -- the outflow is expected to be
{\em clumpy}, rather than smooth, and probably on a variety of scales. 
A clumpy outflow offers a possible solution to the so-called `over-ionization problem' in 
quasar and AGN outflows {\bf \citep[e.g.][]{junk1983,weymann1985,hamann2013}}. 
This is the main motivation for incorporating clumping into our model.

{\bf Deciding on how to implement clumping into our existing wind models was not straightforward.
First, and most importantly, the physical scale lengths and density contrasts in AGN outflows are not well-constrained from observations or theory.  As a result, while one could envision in principle, clouds with a variety of sizes and density contrasts varying perhaps as function of radius, there would have been very little guidance on how to set nominal values of the various parameters of such a model.}  
Second, there are significant computational difficulties associated with adequately resolving and realistically modelling a series of small scale, high density regions with a MCRT
{\bf -- or for that matter, a hydrodynamical --} code. 
Given the lack of knowledge about the actual type of clumping, we have adopted
a simple approximation used successfully in stellar wind modelling, known as 
{\em microclumping} \citep[e.g.][]{hamann1998,hilliermiller1999,hamann2008}.

The underlying assumption of microclumping is that clump sizes are much smaller than the 
typical photon mean free path, and thus the clumps are 
both geometrically and optically thin. {\bf This approach 
allows one to treat clumps only in terms of their volume filling factor, $f_V$, 
instead of having to specify separately their size and density distributions.
In our model, $f_V$ is independent of position (see section~4.4).
The inter-clump medium is assumed to be unimportant in determining the emergent
spectrum, and is therefore modeled as a vacuum,
although the outflow is still non-porous and axisymmetric.}
The density of the clumps is multiplied by the ``density enhancement'' 
$D=1/f_V$. Opacities, $\kappa$, and emissivities, $\epsilon$, 
can then be expressed as 
\begin{equation}
\kappa = f_V \kappa_C(D);~~\epsilon = f_V \epsilon_C(D).
\end{equation}
Here the subscript $C$ denotes that the quantity is calculated using the 
enhanced density in the clump. The resultant effect is that, {\em for fixed temperature},
processes that are linear in density, such as electron scattering, are unchanged, 
as $f_V$ and $D$ will cancel out. However, any quantity that scales with the square of density, 
such as collisional excitation or recombination, will increase by a factor of $D$.
In our models, the temperature is not fixed, and is instead set by balancing heating and 
cooling in a given cell. In the presence of an X-ray source, this thermal balance is 
generally dominated by bound-free heating and line cooling. The main effect of including 
clumping in our modelling is that it moderates the ionization state due to the increased 
density. This allows an increase in the ionizing luminosity, amplifying the amount of
bound-free heating and also increasing the competing line cooling term
(thermal line emission).

{\bf 
The smallest scale length in a MCRT model such as this is normally the Sobolev length, given by
\begin{equation}
l_S = \frac{v_{th}}{| dv/ds |}
\end{equation}
This is typically $\sim10^{13}$~cm near the disc plane, increasing outwards.
We use the mean density to calculate the Sobolev optical depth, which assumes that
$l_S$ is greater than the typical clump size.
Thus for the microclumping assumption to be formally correct, 
clumps should be no larger than $\sim10^{12}$~cm.
This size scale is not unreasonable for quasar outflows, as
\cite{dekool1995} suggest that BAL flows may have low filling factors with
clump sizes of $\sim10^{11}$~cm. }
% An additional concern with this approach to clumping is that the clumps
% may rapidly cool and expand into the inter-clump medium. This is a well-known
% issue in BLR and outflow models (REFs), and requires some kind of confining mechanism,
% such as magnetic confinement \cite[e.g.][]{dekool1995}. Alternatively, as already discussed,
% the LDI may cause clumpy flows as expected in stellar winds.}

Our clumping treatment is necessarily simple; it does not adequately
represent the complex substructures and stratifications in ionization
state we expect in AGN outflows. 
Nevertheless, this parameterization 
allows simple estimates of the effect clumping has on the ionization 
state and emergent line emission.

@@

{\bf The $89^\circ$ viewing angle cannot represent a typical quasar within a unified model, as
extreme inclinations should be extremely underrepresented in quasar samples.
This is in part due to the compton-thick absorber or `torus' 
expected at high inclinations \cite[e.g.][]{antonucci1985,martinez2007}.
However, this spectrum does show that a wind model can
naturally reproduce quasar emission lines if the emissivity of the wind is 
increased {\em with respect to the disc continuum}.
In addition, it neatly demonstrates how a stratified outflow can naturally
reproduce the range of ionization states seen in quasars.}

@@

At high inclinations, the synthetic spectra exhibit blue-shifted BALs in \al\ and \mg --
the absorption lines seen in LoBALQSOs, and we even see absorption in Fe~\textsc{ii}
at the highest inclinations. Line profiles in velocity space 
for \civ, \al\ and \mg, are shown in Fig.~\ref{fig:lobal} for a range
of BALQSO viewing angles. We find that ionization stratification
of the wind causes lower ionization material to have a smaller covering factor, 
as demonstrated by figures~\ref{fig:wind} and \ref{fig:lobal}.
This confirms the behaviour expected from a unification model such as Elvis (2000). 
LoBALs are only present at viewing angles close to edge-on ($i>75^\circ$),
as predicted by polarisation results \citep{brotherton1997}.
As observed in a BALQSO sample by \cite{filizak2014}, we find that
BAL troughs are wider and deeper when low ionization absorption features are present,
and high ionization lines have higher blue-edge velocities than the 
low ionization species.

There is also a correlation between the strength of LoBAL features
and the amount of continuum attenuation at that sightline, particularly
blueward of the Lyman edge as the low ionization base 
intersects the line-of-sight. 
A model such as this therefore predicts that LoBALQSOs and FeLoBALQSOs 
have stronger Lyman edge absorption and 
are more Compton-thick than HiBALQSOs and Type 1 quasars.
An edge-on scenario also offers a potential explanation for the rarity of LoBAL and
FeLoBAL quasars, due to a foreshortened and attenuated continuum, 
although BAL fraction inferences are fraught with complex selection 
effects \citep{goodrich1997,krolikvoit1998}.

@@

Having selected an individual fiducial model from the simulation grid, it is important
to briefly explore how specialised this model is, and how small parameter
changes can affect the synthetic spectra. {\bf Fig.~\ref{fig:grid}
shows the EW for} \civline\ {\bf and} \mgline\ {\bf at a low inclination, 
and $BI$ for} \civline\ {\bf at a high inclination for the simulation
grid. We find that almost all the models with $f_V=0.1$ are over-ionized, and 
fail to produce strong} \civ\ {\bf BALs or emission lines. However, 
the models with $f_V=0.01$ generally produce} \civ\ {\bf BALs and emission lines.
The fiducial model is representative of this family of models and 
the spectra are generally similar, although it is clear from the figure
that we have selected a model near the upper end of the EW distributions.

We find that it is difficult to significantly increase line emission while
keeping the luminosity and mass loss rate of the system fixed.}
We show an additional point on figure 7 corresponding to a model with an order of
magnitude higher X-ray luminosity and double the mass loss rate. As expected, 
this results in far higher line EWs, but fails to produce BALs because
the collisionally excited emission swamps the BAL profile. In addition,
this model would lie well above the expected $L_{2kev}-L_{2500}$ 
relation in figure 5. Such a high X-ray luminosity could therefore 
not be the cause of the strong line emission seen in {\em all} Type 1 quasars.

@@

We have carried out MCRT simulations using a simple
prescription for a biconical disc wind, with
the aim of expanding on the work of H13. To do this, we introduced two main
improvements: First, we included a simple treatment of clumping, and second, 
we improved the modelling of recombination lines by treating H and He as
`macro-atoms'. Having selected a fiducial model from an initial simulation grid,
we assessed the viability of such a model for geometric 
unification of quasars, and found the following main points:
\begin{enumerate}
\item Clumping the wind with a volume filling factor of $0.01$ 
moderates the ionization state
sufficiently to allow for the 
formation of strong UV BALs while agreeing well with the X-ray
properties of luminous AGN and quasars. 
\smallskip
\item A clumpy outflow model naturally 
reproduces the range of ionization states
expected in quasars, due to its stratified density
and temperature structure. 
LoBAL line profiles are seen at a subset of viewing angles, and Fe~\textsc{ii}
absorption is seen at particularly high inclinations. 
\smallskip
\item The synthetic spectra show \la\ line and weak He~\textsc{ii}~$1640$\AA\ line
as a result of our improved treatment of recombination using macro-atoms. We also see
Balmer emission lines and a Balmer recombination continuum in the optical spectrum, but this
is only really significant at high inclination where the continuum is suppressed.  
\smallskip
\item The higher X-ray luminosity causes a significant 
increase in the strength of the collisionally excited emission
lines produced by the model. 
However, the equivalent-width ratios of the emission lines do not match
observations, suggesting that a greater volume of dense ($n_e\sim10^{10}$~cm$^{-3}$)
material may be required.
\smallskip
\item The line EWs in the synthetic spectra increase with inclination.
BAL and non-BAL quasar composites have comparable EWs, so our model
fails to reproduce this behaviour.
 If the BLR emits fairly isotropically then for a 
foreshortened, limb-darkened accretion disc 
it is not possible to achieve line ratios at low inclinations 
that are comparable to those at high inclinations. 
We suggest that understanding the angular distribution of 
line and continuum emission is a crucial question for theoretical models.
\end{enumerate}
Our work confirms a number of expected outcomes from a geometric unification 
model, and suggests that a simple biconical geometry such as this can come close to 
explaining much of the  phenomenology of quasars. However, our conclusions pose 
some challenges to a picture in which BALQSOs are
explained by an {\em equatorial} wind rising from a classical thin disc, and suggest 
the angular distribution of emission is important to understand if this 
geometry is to be refuted or confirmed. We suggest that obtaining reliable 
observational orientation indicators and 
exploring a wider parameter space of outflow geometries in simulations
are obvious avenues for future work.

@@

The work of JHM, SWM, NSH and CK is supported by the
Science and Technology Facilities Council (STFC),
via two studentships and a consolidated grant, respectively.
CK also acknowledges a Leverhulme fellowship.
{\bf We would like to thank the anonymous referee for a helpful
and constructive report.}
We would like to thank Omer Blaes, Ivan Hubeny and Shane Davis for their
assistance with \agn. We are grateful to Mike Brotherton, Mike DiPompeo,
Sebastien Hoenig and Frederic Marin for helpful correspondence regarding
polarisation measurements and orientation indicators.
We would also like to thank Daniel Proga, Daniel Capellupo, Sam Connolly and
Dirk Grupe for useful discussions.  Simulations were conducted using \py\ version 80,
and made use of the IRIDIS High Performance Computing Facility at the
University of Southampton. Figures were produced using the {\tt matplotlib} 
plotting library \citep{matplotlib}. This work made use of the Sloan Digital Sky Survey.
Funding for the Sloan Digital Sky Survey has been provided by
the Alfred P. Sloan Foundation, the U.S. Department of Energy Office of
Science, and the Participating Institutions.
